Robust Bayesian Analysis of Causal Inference Problems

### 2024-03-21 JASA-template.tex
Updates.

%%% MT to TB: there is a slight disjoint between this section and what
%%% comes later: only \delta_i is ever being referred to, \hat{\delta}
%%% is never referred to later on (also I find its definition
%%% confusing); can we simply say that we try to estimate
%%% $\delta_{n+1}$, the main difficulty being that $\delta_1$, \dots,
%%% $\delta_n$ are not directly observable, instead we only observe
%%% each $Y_i$ under either $T_i=0$ or $T_i=1$ but never both?

Let an observational study on $n$ individuals give us the outcomes $Y\coloneqq(Y_1, \dots, Y_n)$ along with 
corresponding treatment indicators $T\coloneqq(T_1, \dots, T_n)$. Then the
treatment 
effect in the population is given by the expectation of the difference
in outcomes between the treatment and controls:
%%% MT to TB: I don't understand this equation; since Y and T are vectors, is \delta simply \delta_1,\dots,\delta_n? I think something else is meant? or remove this equation?
\begin{align}
	\delta = \mathbb{E}(Y\mid T =1) - \mathbb{E}(Y\mid T=0).
\end{align}
Similarly, the individual causal
effect of treatment $T_i$ on outcome $Y_i$ is given by:
\begin{align}
	\delta_i \coloneqq \mathbb{E}(Y_i\mid T_i=1) - \mathbb{E}(Y_i\mid T_i=0).
\end{align}
%%% MT to TB: are we interested in predicting the effect on an individual that is not yet observed, i.e. say $\delta_{n+1}$?
That is, we are interested in the difference between the outcomes when the $i$-th subject receives
the treatment and when they remain as a control.

In theory, both $\mathbb{E}(Y_i\mid T_i=1)$ and $\mathbb{E}(Y_i\mid T_i=0)$ exist.
However, we cannot observe them simultaneously, since we cannot `treat' and `not treat' the same individual at the same time.
What we can observe is the average causal effect of the treatment,
by calculating the averaged
outcome of all the subjects that received the treatment
minus that of all the subjects that remained as control:
\begin{align}
  \hat{\delta} \coloneqq 
  %\frac{\sum_{i=1}^n Y_i\cdot\mathbb{I}(T_i=1)-\sum_{i=1}^n Y_i\cdot\mathbb{I}(T_i=0)}{n}
  %%% MT to TB check if this is correct, it seems a bit simpler to write like this
  \frac{\sum_{i=1}^n Y_iT_i-\sum_{i=1}^n Y_i(1-T_i)}{n}.
\end{align}
%%% MT to TB: is it assumed both groups have the same size, namely n/2? or not?
However, this relies on an important assumption that the treatment effect
on the $i$-th subject given that they received the treatment is the same as the (counterfactual) treatment effect when they remain as control \cite{winship99}.

\subsection{Regression Model}
Regression methods are widely used in causal effect estimation. The
main idea behind these regression methods is to remove the
correlation between the treatment indicator and the error term
\cite{winship99,HECKMAN1985}. To do so, we rely on $p$ different observed quantities
or predictors denoted by $X\coloneqq$ $[X_1^T, \dots, X_n^T]^T$
where each $X_i\in\mathbb{R}^p$.
Each $X_i$ is treated as a $p$-dimensional row vector,
so $X$ is a $n\times p$ matrix.
Now, let
$\beta \coloneqq (\beta_1$, \dots, $\beta_p)^T$ denote the vector of regression
coefficients
related to the predictors, and $\beta_T$ denote a regression coefficient related to the treatment.
Then we can define a linear model for the outcome
so that
\begin{equation}
	Y_i =  T_i \beta_{T} + X_i\beta + \epsilon_i
\end{equation}
where $\epsilon_i\sim \mathcal{N}(0, \sigma^2)$. Clearly, when
the underlying true outcome model is linear with respect to the treatment,
%%% MT to BT: should we say $\delta_{n+1}$ here (and Y_{n+1}, T_{n+1})?
\begin{equation}
	\delta_i = \mathbb{E}(Y_i\mid T_i =1) - \mathbb{E}(Y_i\mid T_i=0) = \beta_{T}.
\end{equation}
*** MT to TB: trying to derive this\dots ***      
\begin{align}
  \delta_i
  &=\mathbb{E}(Y_i\mid T_i =1) - \mathbb{E}(Y_i\mid T_i=0)\\
  &=\beta_T+\mathbb{E}(X_i\beta+\epsilon_i\mid T_i =1) - \mathbb{E}(X_i\beta+\epsilon_i\mid T_i=0)\\
  %%% MT to BT:
  \intertext{*** I get why $\epsilon_i$ has expectation zero (assumed independent from $T_i$); but why do the $X_i\beta$ terms vanish, since $X_i$ and $T_i$ are not assumed independent? maybe this can only assumed to hold for $i=n+1$ i.e. for the `next' person, not for the person in the data\dots another even better solution is just to condition on $X_{n+1}$ since this is known, then the entire problem goes away ***}
  &=\beta_{T}.
\end{align}
*** MT to BT: derivation v2; this shows the causal effect does not depend on the value of $X_{n+1}$: ***
\begin{align}
  \delta(x_{n+1})
  &=\mathbb{E}(Y_{n+1}\mid X_{n+1}=x_{n+1},T_{n+1} =1) - \mathbb{E}(Y_{n+1}\mid X_{n+1}=x_{n+1},T_{n+1}=0)\\
  &=\beta_T+x_{n+1}\beta+\mathbb{E}(\epsilon_{n+1}\mid X_{n+1}=x_{n+1},T_{n+1} =1) \\
  &\quad - x_{n+1}\beta-\mathbb{E}(\epsilon_{n+1}\mid X_{n+1}=x_{n+1},T_{n+1} =0)\\
  %%% MT to BT:
  \intertext{and because $\epsilon_{n+1}$ is independent from $X_{n+1}$ and $T_{n+1}$,}
  &=\beta_{T}.
\end{align}
Note that, for this model, the causal effect $\delta(x_{n+1})$ does not depend on the value $x_{n+1}$ of $X_{n+1}$.

>>>>

\subsection{Regression Model}

Let an observational study on $n$ individuals give us
the outcomes $Y\coloneqq(Y_1, \dots, Y_n)$ along with 
corresponding treatment indicators $T\coloneqq(T_1, \dots, T_n)$.
Regression methods are widely used in causal effect estimation. The
main idea behind these regression methods is to remove the
correlation between the treatment indicator and the error term
\cite{winship99,HECKMAN1985}.

%%% MT to TB: not sure where this text needs to go, omitted for now...

% In theory, both $\mathbb{E}(Y_i\mid T_i=1)$ and $\mathbb{E}(Y_i\mid T_i=0)$ exist.
% However, we cannot observe them simultaneously, since we cannot `treat' and `not treat' the same individual at the same time.

% We rely on the important assumption that the treatment effect
% on the $i$-th subject given that they received the treatment is the same as the (counterfactual) treatment effect when they remain as control \cite{winship99}.

To do so, we rely on $p$ different observed quantities
or predictors denoted by $X\coloneqq$ $[X_1^T, \dots, X_n^T]^T$
where each $X_i\in\mathbb{R}^p$.
Each $X_i$ is treated as a $p$-dimensional row vector,
so $X$ is a $n\times p$ matrix.
Now, let
$\beta \coloneqq (\beta_1$, \dots, $\beta_p)^T$ denote the vector of regression
coefficients
related to the predictors, and $\beta_T$ denote a regression coefficient related to the treatment.
Then we can define a linear model for the outcome
so that
\begin{equation}
	Y_i =  T_i \beta_{T} + X_i\beta + \epsilon_i
\end{equation}
where $\epsilon_i\sim \mathcal{N}(0, \sigma^2)$.

%%% MT to TB: check if you're happy with this paragraph
To decide whether or not to treat a new individual with given predictors,
we are mainly interested in the effect of the treatment on the outcome.
More precisely, the treatment effect of a new individual, indexed as $n+1$,
whose outcome $Y_{n+1}$ is not yet observed, and with observed predictors $X_{n+1}=x_{n+1}$, is defined by:
\begin{align}
  \delta(x_{n+1})
  &\coloneqq\mathbb{E}(Y_{n+1}\mid X_{n+1}=x_{n+1},T_{n+1} =1) - \mathbb{E}(Y_{n+1}\mid X_{n+1}=x_{n+1},T_{n+1}=0)
\end{align}
For our model, due to linearity of expectation, we have that
\begin{align}
  \delta(x_{n+1})
  &=\beta_T+x_{n+1}\beta+\mathbb{E}(\epsilon_{n+1}\mid X_{n+1}=x_{n+1},T_{n+1} =1) \\
  &\quad - x_{n+1}\beta-\mathbb{E}(\epsilon_{n+1}\mid X_{n+1}=x_{n+1},T_{n+1} =0)\\
  \intertext{and because $\epsilon_{n+1}$ is independent from $X_{n+1}$ and $T_{n+1}$,}
  &=\beta_{T}.
\end{align}
Note that, for this model, the treatment effect $\delta(x_{n+1})$ does not depend on the value $x_{n+1}$ of $X_{n+1}$.
So, to find the treatment effect, we simply need to estimate $\beta_T$.
